Estimating fitness costs from mutation counts

### 2023-01-05 notes/fitness_inference.tex
add notes
Richard's notes on calculations (#10)
* add notes

* add notes

* add notes

Co-authored-by: Richard Neher <richard.neher@unibas.ch>

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}
\begin{document}
\title{Estimating fitness costs from mutation counts}
\author{Richard Neher}
\maketitle

@@

At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
\begin{equation}
    \begin{split}
        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
        &\approx \begin{cases}
            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
        \end{cases}
    \end{split}
\end{equation}
This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
In this case, the above simplifies to
\begin{equation}
    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
\end{equation}
This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
In growing population, the approximation fails if the product of fitness effect and the time over which the variant is grown is large, i.e., if the fitness effect is substantial on this time scale.
Outside this parameter range, there is still a dependence on $s$, but it is weaker.

>>>>

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}
\begin{document}
\title{Estimating fitness costs from mutation counts}
\maketitle

@@

At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
\begin{equation}
    \begin{split}
        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
        &\approx \begin{cases}
            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
        \end{cases}
    \end{split}
\end{equation}
This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
In this case, the above simplifies to
\begin{equation}
    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
\end{equation}

In a variant that has been growing with rate $\gamma_0$ for a time $\tau=T-t$ and sampled with $z=\epsilon$, we thus expect that the number of times we observe separate mutant lineages depends on $s$ as
\begin{equation}
    \langle \phi \rangle \approx \frac{\epsilon}{\gamma_0(\epsilon+s)}
\end{equation}
This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is large, i.e., if the fitness effect is substantial on this time scale.
Outside this parameter range, there is still a dependence on $s$, but it is weaker.
